MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-20 Manuscript/Manuscript-1-DCA.tex
vsn 1

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. First
we use computer simulations across differing numbers of species,
differently shaped phylogenies, and differing input levels of
phylogenetic signal, to explore the degree to which \(\lambda\)
correctly identifies known levels of phylogenetic signal, and under what
circumstances. For comparison, we also evaluate estimates of \(Kappa\)
and its empirically-derived effect size (\(Z_K\)) for this purpose.
Additionally, we use simulations to determine how the inclusion of
\(\lambda\) in phylogenetic regression and ANOVA (i.e., PGLS) affects
parameter estimation, and whether levels of phylogenetic signal
estimated in a PGLS framework are accurate. We then survey the recent
macroevolutionary literature for published papers containing estimates
of \(\lambda\) from empirical datasets, and compare these empirical
estimates to patterns gleaned from our computer simulations. In general
we find that while PGLS parameters (e.g., \(\beta\)) are accurately
estimated with the inclusion of phylogenetic signal, estimates of
\(\lambda\) are not. Further, we find that estimates of \(\lambda\) vary
widely for a given input value of phylogenetic signal, and that this
variation is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. By contrast, variation across effect sizes (\(Z_K\)) obtained
from \(Kappa\) is far more consistent across the range of input values,
and that the precision of estimating \(Z_K\) is preferable to that
observed using \(\lambda\). We conclude that estimates of phylogenetic
signal using Pagel's \(\lambda\) are often innacurate, and thus
interpreting strength of phylogenetic signal in phenotypic datasets
based on this measure should be treated with caution. By contrast,
effect sizes obtained from \(Kappa\) hold promise for characterizing
phylogenetic signal, and comparing the strength of phylogenetic signal
across datasets.

@@

We conducted a series of computer simulations to evaluate the ability of
Pagel's (1999) \(\lambda\) to estimate known levels of phylogenetic
signal. Our primary simulations were based on pure-birth phylogenies;
however, we also evaluated patterns on both balanced and pectinate trees
to determine whether tree shape affected our findings (see Supporting
Information). Our first set of simulations evaluated the extent to which
values of \(\lambda\) estimated from the data corresponded with actual
input levels of phylogenetic signal. For these simulations we generated
50 pure-birth phylogenies at each of six different tree sizes, ranging
from 32 to 1024 taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the
simulated phylogenies by multiplying the internal branches by
\(\lambda\). We used a set of values that encompassed the entire range
of \(\lambda\): (\(\lambda_{in} = 0.0 \to 1.0\); in 21 intervals of 0.05
units), resulting in 1050 scaled phylogenies at each level of species
richness (\(n\)). Continuous traits were then simulated on each
phylogeny under a Brownian motion model of evolution to obtain datasets
with known and differing levels of phylogenetic signal. These varied
from datasets with no phylogenetic signal (when \(\lambda_{in} =0\)) to
datsets with phylogenetic signal corresponding to what was expected
under Brownian motion (when \(\lambda_{in} =1\)). \hfill\break

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda_{est}\) was approximated by
observing the variation of \(\lambda_{est}\) obtained at each input
level of phylogenetic signal (\(\lambda_{in}\)). Finally, for comparison
we estimated the degree of phylogenetic signal in each dataset using
\(Kappa\) (Blomberg et al. 2003; Adams 2014a), and from this calculated
its effect size (\(Z_K\): sensu Adams and Collyer 2016, 2019a) for each
dataset. Here, the effect size was estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution (for discussion see: Adams
and Collyer 2019b). Because \(Z_K\) differs in scale from \(\lambda\),
we used a linear normalization to standardize \(Z_K\) to a uniform
distribution (\(0\rightarrow1\)), and variation in the set of normalized
\(Z_K\) values at each input level of phylogenetic signal
(\(\lambda_{in}\)) were then estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of
\(\lambda\) estimated in PGLS regression and ANOVA (i.e., \(Y\sim{X}\))
corresponded with actual input levels of phylogenetic signal in the
response variable. As before we generated 50 pure-birth phylogenies at
each of six levels of species richness (\(n=2^5 - 2^{10}\)), and
rescaled each with a set of input \(\lambda\) values
(\(\lambda_{in} = 0.0 - 1.0\); in 21 intervals of 0.05 units). Next, an
independent variable \(X\) was simulated on each phylogeny under a
Brownian motion model of evolution (for PGLS regression). For
phylogenetic ANOVA, random groups (\(X\)) were obtained by simulating a
discrete (binary) character on each phylogeny. Next, the dependent
variable was simulated in such a manner as to contain a known
relationship with \(X\) plus random error containing phylogenetic
signal. This was accomplished as: \(Y=\beta{X}+\epsilon\). Here, the
association between \(Y\) and \(X\) was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). For each dataset, the fit
of the phylogenetic regression was estimated using maximum likelihood,
and parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained and evaluated as above. All analyses were performed in R v3.6.0
(R Core Team 2019) using the packages \texttt{geiger} (Harmon et al.
2008), \texttt{caper} (Orme et al. 2013), \texttt{phytools} (Revell
2012), and \texttt{geomorph} (Adams and Otárola-Castillo 2013; Adams et
al. 2020). R-scripts are found in the Supplemental Material.

@@

lambda: (Fig 1) not good predictor of input physignal (regression),
particularly at low sample sizes. Also, similar to early results of
Boettiger (but greatly expanded) range and variation of lambda very
large at low N, and generally decreases with larger N. However, the
precision was not uniform across the range of lambda-in. Variation in
lambda-est was greatest at intermediate levels of physig, and descreased
at lower and higher values. Thus, the overall dispersion across the
range of input values resembled an asymmetric lens (\textbf{NAME?}),
with less variation at lambda-in close to 1, greatest variation in
middle, and less (but substantial) variation near lambda-in of zero.
Notably, even at very large species richness (n=1024) the range of
lam-est was quite large: e.g., l-in 0.5 has range of 45\% of all
possible values. Thus, lambda-est is not very precise in terms of
estimating the true l-in parameter, meaning that using lambda to
biologically interpret the `strength' of phylogenetic signal in a
dataset, or compare between datasets, is fraught with difficulty.
\hfill\break

@@

2: expand on Lambda.. lambda innacurate, not precise, level of precision
varies with input physig (worse in mid-range). Patterns worse with PGLS,
though beta still estimated properly. Conclusion, lambda not overly
useful.

@@

\textbf{Figure 2}. Variation in estimates of phylogenetic signal across
input levels of phylogenetic signal. \textbf{A}: Estimates of Pagel's
\(\lambda\) for data simulated on phylogenies with 128 taxa (\(n=128\)),
\textbf{B} Estimates of \(Z_K\) for data simulated on phylogenies with
128 taxa (\(n=128\)), \textbf{C} Variance in the variation of
\(\lambda_{est}\) across input levels of phylogenetic signal, estimated
on phylogenies containing differing numbers of species. \textbf{D}
Variance in the variation of \(Z_K\) across input levels of phylogenetic
signal, estimated on phylogenies containing differing numbers of
species. \hfill\break

@@

\singlespacing \textbf{Figure 2}. Variation in estimates of phylogenetic
signal across input levels of phylogenetic signal. \textbf{A}: Estimates
of Pagel's \(\lambda\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), \textbf{B} Estimates of \(Z_K\) for data simulated on
phylogenies with 128 taxa (\(n=128\)), \textbf{C} Variance in the
variation of \(\lambda_{est}\) across input levels of phylogenetic
signal, estimated on phylogenies containing differing numbers of
species. \textbf{D} Variance in the variation of \(Z_K\) across input
levels of phylogenetic signal, estimated on phylogenies containing
differing numbers of species.

@@

>>>>

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. First
we use computer simulations across differing numbers of species,
differently shaped phylogenies, and differing input levels of
phylogenetic signal, to explore the degree to which \(\lambda\)
correctly identifies known levels of phylogenetic signal, and under what
circumstances. For comparison, we also evaluate estimates of \(Kappa\)
and its empirically-derived effect size (\(Z_K\)) for this purpose.
Additionally, we use simulations to determine how the inclusion of
\(\lambda\) in phylogenetic regression and ANOVA (i.e., PGLS) affects
parameter estimation, and whether levels of phylogenetic signal
estimated in a PGLS framework are accurate. We then survey the recent
macroevolutionary literature for published papers containing estimates
of \(\lambda\) from empirical datasets, and compare these empirical
estimates to patterns gleaned from our computer simulations. In general
we find that while PGLS parameters (e.g., \(\beta\)) are accurately
estimated with the inclusion of phylogenetic signal, estimates of
\(\lambda\) are not. Further, we find that estimates of \(\lambda\) vary
widely for a given input value of phylogenetic signal, and that this
variation is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. By contrast, variation across effect sizes (\(Z_K\)) obtained
from \(Kappa\) is far more consistent across the range of input values,
and the precision of estimating \(Z_K\) is preferable to that observed
using \(\lambda\). We conclude that estimates of phylogenetic signal
using Pagel's \(\lambda\) are often innacurate, and thus interpreting
strength of phylogenetic signal in phenotypic datasets based on this
measure should be treated with caution. By contrast, effect sizes
obtained from \(Kappa\) hold promise for characterizing phylogenetic
signal, and comparing the strength of phylogenetic signal across
datasets.

@@

We conducted a series of computer simulations to evaluate the ability of
Pagel's \(\lambda\) to estimate known levels of phylogenetic signal. Our
primary simulations were based on pure-birth phylogenies; however, we
also evaluated patterns on both balanced and pectinate trees to
determine whether tree shape affected our findings (see Supporting
Information). Our first set of simulations evaluated the extent to which
values of \(\lambda\) estimated from the data corresponded with actual
input levels of phylogenetic signal. For these simulations we generated
50 pure-birth phylogenies at each of six different tree sizes, ranging
from 32 to 1024 taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the
simulated phylogenies by multiplying the internal branches by
\(\lambda\). We used a set of values that encompassed the entire range
of \(\lambda\): (\(\lambda_{in} = 0.0 \to 1.0\); in 21 intervals of 0.05
units), resulting in 1050 scaled phylogenies at each level of species
richness (\(n\)). Continuous traits were then simulated on each
phylogeny under a Brownian motion model of evolution to obtain datasets
with known and differing levels of phylogenetic signal. These varied
from datasets with no phylogenetic signal (when \(\lambda_{in} =0\)) to
datsets with phylogenetic signal corresponding to what was expected
under Brownian motion (when \(\lambda_{in} =1\)). \hfill\break

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda\) was approximated by observing
the variation of \(\lambda_{est}\) obtained at each input level of
phylogenetic signal (\(\lambda_{in}\)). Finally, for comparison we
estimated the degree of phylogenetic signal in each dataset using
\(Kappa\) (Blomberg et al. 2003; Adams 2014a), and from this calculated
its effect size (\(Z_K\): sensu Adams and Collyer 2016, 2019a) for each
dataset. Here, the effect size was estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution (for discussion see: Adams
and Collyer 2019b). Because \(Z_K\) differs in scale from \(\lambda\),
we used a linear normalization to standardize \(Z_K\) to a uniform
distribution (\(0\rightarrow1\)), and variation in the set of normalized
\(Z_K\) values at each input level of phylogenetic signal
(\(\lambda_{in}\)) were then estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of
\(\lambda\) estimated in PGLS regression and ANOVA (i.e., \(Y\sim{X}\))
corresponded with actual input levels of phylogenetic signal in the
response variable. As before we generated 50 pure-birth phylogenies at
each of six levels of species richness (\(n=2^5 - 2^{10}\)), and
rescaled each with a set of input \(\lambda\) values
(\(\lambda_{in} = 0.0 - 1.0\); in 21 intervals of 0.05 units). Next, an
independent variable \(X\) was simulated on each phylogeny under a
Brownian motion model of evolution (for PGLS regression). For
phylogenetic ANOVA, random groups (\(X\)) were obtained by simulating a
discrete (binary) character on each phylogeny. Next, the dependent
variable was simulated in such a manner as to contain a known
relationship with \(X\) plus random error containing phylogenetic
signal. This was accomplished as: \(Y=\beta{X}+\epsilon\). Here, the
association between \(Y\) and \(X\) was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). For each dataset, the fit
of the phylogenetic regression was estimated using maximum likelihood,
and parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained and evaluated as above. All analyses were performed in R v3.6.0
(R Core Team 2019) using the packages \texttt{geiger} (Harmon et al.
2008), \texttt{caper} (Orme et al. 2013), \texttt{phytools} (Revell
2012), and \texttt{geomorph} (Adams and Otárola-Castillo 2013; Adams et
al. 2020). R-scripts are found in the Supporting Information.

@@

Our primary series of simulations revealed several patterns. First, the
relationship between \(\lambda_{est}\) and input levels of phylogenetic
signal was \(\sim\beta=1.0\) at each level of species richness
(\(n=32, 64, 128, 256, 512, 1024\)). This implied that the mean
\(\lambda_{est}\) across datasets for a given set of simulation
conditions reflected input levels of phylogenetic signal. However, as
shown in Figure 1, the precision of those estimates varied widely, and
in several interesting ways. Predictably, precision was worst at low
levels of species richness, where in many cases the set of
\(\lambda_{est}\) spanned nearly the entire range of possible values
(e.g., \(n=32\); \(\lambda_{est}=0.5\): range = \(0.0\to0.985\)).
Second, as species richness increased, variation across estimates of
\(\lambda\) decreased (Figure 1). This confirmed the pattern identified
by Boettiger et al.~(2012) on a small (\(n=13\)) and large (\(n=281\))
phylogeny, and demonstrated that this trend was general across a wider
range of species richness. Importantly however, our broader set of
simulations revealed that the precision of \(\lambda_{est}\) was not
uniform across all levels of phylogenetic signal, and that there was a
particular pattern to the imprecision of estimating this phylogenetic
parameter. Specifically, we found that variation in \(\lambda_{est}\)
was highest at intermediate levels of phylogenetic signal, and decreased
at both lower and higher levels of input signal (Figure 1). This implied
that the precision in estimating \(\lambda\) was worst at intermediate
levels of phylogenetic signal; a pattern that remained consistent across
all levels of species richness evaluated here (Figure 1). Notably, even
at very large levels of species richness, the range of \(\lambda_{est}\)
still encompassed a substantial portion of possible values (e.g.,
\(n=1024\); \(\lambda_{est}=0.5\): range = \(0.31\to 0.68\)). From this
we conclude that \(\lambda_{est}\) is not an accurate estimator of the
strength of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the level of phylogenetic signal based on
\(\lambda\) should be treated with caution. \hfill\break

@@

2: expand on Lambda.. lambda innacurate, not precise, level of precision
varies with input physig (worse in mid-range). NEW RESULT. We are first
to show this. NOTE: pattern is obvious with reflection. Since it is a
`bounded' parameter estimation should be best at the extremes\ldots{}
(state this?).. hmm.

Patterns worse with PGLS, though beta still estimated properly.
Conclusion, lambda not overly useful.

@@

\textbf{Figure 2}. Variation in estimates of phylogenetic signal across
input levels of phylogenetic signal. (A) Estimates of Pagel's
\(\lambda\) for data simulated on phylogenies with 128 taxa (\(n=128\)),
(B) Estimates of \(Z_K\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (C) Variance in the variation of \(\lambda_{est}\) across
input levels of phylogenetic signal, estimated on phylogenies containing
differing numbers of species. (D) Variance in the variation of \(Z_K\)
across input levels of phylogenetic signal, estimated on phylogenies
containing differing numbers of species. \hfill\break

@@

\singlespacing \textbf{Figure 2}. Variation in estimates of phylogenetic
signal across input levels of phylogenetic signal. (A) Estimates of
Pagel's \(\lambda\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (B) Estimates of \(Z_K\) for data simulated on phylogenies
with 128 taxa (\(n=128\)), (C) Variance in the variation of
\(\lambda_{est}\) across input levels of phylogenetic signal, estimated
on phylogenies containing differing numbers of species. (D) Variance in
the variation of \(Z_K\) across input levels of phylogenetic signal,
estimated on phylogenies containing differing numbers of species.

@@

\includegraphics[width=0.95\linewidth]{Fig3}
